### 2017-01-26 kalman_walk.tex
josh edits to intro

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[hidelinks]{hyperref}

@@

% Outline:
% 
% - What the problems are:
% 
%     * evolution of perfectly fit populations
% 
% - Motivations: Can you infer function from structure? And...
% 
%     * motifs
%     * speciation
%     * network size
%     * rube-goldberginess
%     * redundancy, robustness
%     * evolvability related to minimality?
% 
% - Give a concrete, simple example.
% 
% - Modeling framework (LTI systems)
%
%     * y = "phenotype"
%     * x = "internal state" or "kryptotype"?
%     * A = "regulatory matrix" or "genetic network"?
%     * Kalman decomposition
% 
% - Evolution of free parameters in the Kalman decomposition
%
%     * Brownian motion on free parameters
%     * how mutations map into this space
%     * dimension: duplication, deletion, recruitment of new things
% 
% - (Compute things)
% 
% - (Allow non-fit stuff too.)

\newcommand{\plr}[1]{{\color{blue}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}

@@

@@

Since the system is linear, the state of the system at any time
is the superposition of its responses to all previous inputs.
This implies that 
if $G(t)$ is the transient response of the system to a unit impulse,
$t$ units of time later,
and $x(0) = 0$,
then the time course of the concentration of the thing we care about, $x_1(t)$, 
can be written as
\begin{align} \label{eqn:ex_convolution}
    x_1(t) &= \int_0^t G(t-s) u(s) ds .
\end{align}
However, it turns out that 
\plr{look for simple demontration this is true}
\begin{align*}
    P_{p} = \left[\begin{matrix} 
        1 & 0 \\ 
        p & 1-p 
    \end{matrix}\right],
\end{align*}
then for any $p \neq 1$, 
if we replace the regulatory matrix $A$ with
\begin{align*}
    A(p) = P^{-1} A P,
\end{align*}
then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
(However, $x_2$ may well be different!)

@@

We will now lay out the model in more general terms.
Suppose that the \emph{internal state} of the system
is parameterized by the concentrations of a collection of $n$ molecular species,
$S_1, \ldots, S_n$,
and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
There are also $m$ ``input'' species, whose concentrations are determined
exogenously to the system,
and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
and $\ell$ ``output'' species, whose concentrations are denoted
$y(t) = (y_1(t),\ldots,y_\ell(t))$.
The output is merely a linear function of the internal state:
\begin{align*}
    y_i(t) = \sum_j C_{ij} x_i(t).
\end{align*}
Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
and sometimes in contrast refer to $x$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
\plr{Maybe.}
The rate at which the $i^\text{th}$ species is produced
is a weighted sum of the concentrations of the other species
as well as the input:
\begin{align*}
    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
\end{align*}
In matrix notation, this is written more concisely as
\begin{align*}
    \ddt x(t) &= A x(t) + B u(t) \\
    y(t) &= C x(t) .
\end{align*}

@@

Given an initial condition and an input, 
it is possible to write the solution $x(t)$
as a convolution with an input-response kernel as in the example above
(equation \eqref{eqn:ex_convolution}).
An alternative way of describing the input-output relationship,
more common in engineering,
is instead to find the \emph{transfer function}
\begin{align*}
    H(z) = C(z I - A)^{-1} B,
\end{align*}
which is more mathematically readable \plr{if you know what it means}.
This can be thought of as the system's response to input at ``frequency'' $z$.
The fact that the transfer function uniquely determines the system's
input-output relationship (i.e., the mapping $u \mapsto y$)
follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
is the Laplace transform of $x$,
and $\tilde u(z)$ is the Laplace transform of $u$,
then
\begin{align*}
    \tilde y(z) = H(z) \tilde u(z) .
\end{align*}
(This happens because the Laplace transform takes convolutions to products,
and $H$ is the Laplace transform of the kernel $G$.)

@@

One way to get equivalent things is change of basis (``algebraic equivlance'').
If multiple realizations of a system are algebraically equivalent, then they
must also have identical transfer functions $G(s) = \overline{G}(s)$. However,
the opposite relation is not necessarily true. Multiple systems can have
identical transfer functions and not have algebraic equivlaence. In fact, two
systems with identical transfer functions can even have different dimensions.

But, that's not all: Kalman decomposition.  
Theorem that all equivalent systems are of this form; 
and representation is unique.

@@

>>>>

\documentclass[11 pt]{article}
\usepackage{amsmath, amssymb, color, xcolor}
\usepackage{graphicx}
\usepackage{fullpage}
\usepackage[hidelinks]{hyperref}

@@

% Outline:
% 
% - What the problems are:
% 
%     * evolution of perfectly fit populations
% 
% - Motivations: Can you infer function from structure? And...
% 
%     * motifs
%     * speciation
%     * network size
%     * rube-goldberginess
%     * redundancy, robustness
%     * evolvability related to minimality?
% 
% - Give a concrete, simple example.
% 
% - Modeling framework (LTI systems)
%
%     * y = "phenotype"
%     * x = "internal state" or "kryptotype"?
%     * A = "regulatory matrix" or "genetic network"?
%     * Kalman decomposition
% 
% - Evolution of free parameters in the Kalman decomposition
%
%     * Brownian motion on free parameters
%     * how mutations map into this space
%     * dimension: duplication, deletion, recruitment of new things
% 
% - (Compute things)
%
% - Generalize ideas to nonlinear systems
%
%   * compute isomorphic or dynamically equivallent nonlinear systems of the feed-forward motif or of the three-node adaptation motif. 
% 
% - (Allow non-fit stuff too.)

% is y the phenotype or H(z) the phenotype? Do we need new jargon anyway?

\newcommand{\plr}[1]{{\color{blue}\it #1}}
\newcommand{\jss}[1]{{\color{olive}\it #1}}
% \newcommand{\ddt}{\frac{d}{dt}}
\newcommand{\ddt}{\dot}

@@

\jss{Introduction}

It is commonly taught that an organism's genome contains the heritable material 
that natural selection filters and that an organism's phenotype directly determines its
evolutionary fitness. Between genotype and phenotype is an often complicated and poorly
undrstood molecular machinery -- and it is a major goal common to many different disciplines
within the life sciences to elucidate its form, function, and evolution. These aims are
delicately intertwined and a comphrehensive understanding of a system's evolution requires
an understanding of its function and vice versa.

The molecular machinery, interacting with the environment, and bridging genotype to phenotype
can be mathematically described as a dynamical system (or a system of differential equations).
Thus, an idealized study of evolution would benefit from knowledge of genomes, phenotypes,
and importantly detailed descriptions of genotype-phenotype maps. Movement in this direction
is ongoing, as researhers have begun to study the evolution of empirically inspired
computational and mathematical models of gene regulatory networks (GRNs) and metabolic
networks that incorporate empirical data such as sequences and expression patterns. If we
allow the reasonable assumption that the genotype-phenotype map can be represented as a
system of differential equations, we can immediately discuss its evolution and function
in a much more mechanistic, yet general, manner.

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models can
fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that there
can be uncertaintity about an inferred model's parameters or even its claims about causal
structure, even with access to complete and perfect data. Models with different parameter
schemes, or even different mechanics can be equally accurate, but still not \emph{actually}
agree with what is being modelled. Although this may frusturate the occasional engineer or
scientist, viewed from another angle, the concepts of unidentifiability and
indistinguishability can provide a starting point for thinking about externally equivalent
systems -- systems that evolution can explore, so long as the parameters and structures can
be realized biologically.

The set of all systems -- that is all model structures and all parameter schemes -- that
are externally, or phenomenologically, equivalent, biologically realizable, and mutationally accessible,

What are typically referred to as ``design principles'' or some sort of clear structure-function relationships are commonly saught

Seeking to attribute functionality, uninformed by evolution, can lead to spurious claims
(i.e. see Graur).

@@

Since the system is linear, the state of the system at any time
is the superposition of its responses to all previous inputs.
This implies that 
if $G(t)$ is the transient response of the system to a unit impulse,
$t$ units of time later,
and $x(0) = 0$,
then the time course of the concentration of the thing we care about, $x_1(t)$, 
can be written as
\begin{align} \label{eqn:ex_convolution}
    x_1(t) &= \int_0^t G(t-s) u(s) ds .
\end{align}
However, it turns out that 
\plr{look for simple demontration this is true}
\begin{align*}
    P_{p} = \left[\begin{matrix} 
        1 & 0 \\ 
        p & 1-p 
    \end{matrix}\right],
\end{align*}
then for any $p \neq 1$, 
if we replace the regulatory matrix $A$ with
\begin{align*}
    A(p) = P A P^{-1},
\end{align*}
then the response of $x_1(t)$ to any particular input $u(t)$ is identical to the original system.
(However, $x_2$ may well be different!)

@@

We will now lay out the model in more general terms.
Suppose that the \emph{internal state} of the system
is parameterized by the concentrations of a collection of $n$ molecular species,
$S_1, \ldots, S_n$,
and the vector of concentrations at time $t$ we denote $x(t)=(x_1(t),\ldots,x_n(t))$.
There are also $m$ ``input'' species, whose concentrations are determined
exogenously to the system,
and are denoted $u(t) = (u_1(t),\ldots,u_m(t))$,
and $\ell$ ``output'' species, whose concentrations are denoted
$y(t) = (y_1(t),\ldots,y_\ell(t))$.
The output is merely a linear function of the internal state:
\begin{align*}
    y_i(t) = \sum_j C_{ij} x_i(t).
\end{align*}
Since $y$ is what natural selection acts on, we refer to it as the \emph{phenotype},
and sometimes in contrast refer to $x$ as the \emph{kryptotype},
as it is ``hidden'' from direct selection.
\plr{Maybe.}
The rate at which the $i^\text{th}$ species is produced
is a weighted sum of the concentrations of the other species
as well as the input:
\begin{align*}
    \ddt x_i(t) = \sum_j A_{ij} x_j(t) + \sum_k B_{ik} u_k(t) .
\end{align*}
In matrix notation, this is written more concisely as
\begin{align} \label{eqn:lti_system}
    \ddt x(t) &= A x(t) + B u(t) \\
    y(t) &= C x(t) .
\end{align}

@@

Given an initial condition and an input, 
it is possible to write the solution $x(t)$
as a convolution with an input-response kernel as in the example above
(equation \eqref{eqn:ex_convolution}).
An alternative way of describing the input-output relationship,
more common in engineering,
is instead to find the \emph{transfer function}
\begin{align} \label{eqn:transfer_fn}
    H(z) = C(z I - A)^{-1} B,
\end{align}
which is more mathematically readable \plr{if you know what it means}.
This can be thought of as the system's response to input at ``frequency'' $z$.
The fact that the transfer function uniquely determines the system's
input-output relationship (i.e., the mapping $u \mapsto y$)
follows from the fact that if $\tilde y(z) = \int_0^\infty e^{-zt} y(t) dt$
is the Laplace transform of $x$,
and $\tilde u(z)$ is the Laplace transform of $u$,
then
\begin{align*}
    \tilde y(z) = H(z) \tilde u(z) .
\end{align*}
(This happens because the Laplace transform takes convolutions to products,
and $H$ is the Laplace transform of the kernel $G$.)

@@

\begin{definition}[Phenomenological equivalence of systems]
    Let $(x(t),y(t))$ and $(\bar x(t),\bar y(t))$ be the solutions to \eqref{eqn:lti_system}
    with coefficient matrices $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ respectively,
    and both $x(0)$ and $\bar x(0)$ are zero. 
    The systems defined by $(A,B,C)$ and $(\bar A,\bar B,\bar C)$ are
    \textbf{phenomenologically equivalent} if
    \begin{align*}
        y(t) = \bar y(t) \qquad \text{for all} t \ge 0.
    \end{align*}
    Equivalently, this occurs if and only if
    \begin{align*}
        H(z) = \bar H(z)  \qquad \text{for all} z \ge 0,
    \end{align*}
    where $H$ and $\bar H$ are the transfer functions of the two systems.
\end{definition}

One way to find other systems equvalent to a given one
is by change of coordinates (``algebraic equivalance''):
if $T$ is an invertible matrix, then the systems $(A,B,C)$ and $(TAT^{-1},TB,CT^{-1})$
have the same dyanamics because their transfer functions are equal:
\begin{align*}
    CT^{-1}( zI - TAT^{-1})^{-1}TB
    =
    CT^{-1}T( zI - A)^{-1}T^{-1}TB
    =
    C( zI - A)^{-1}B .
\end{align*}
However, the converse is not necessarily true: 
systems can have identical transfer functions without being changes of coordinates of each other.
In fact, systems with identical transfer functions can involve interactions between different
numbers of molecular species.

The general problem of how to characterize the set of equivalent
But, that's not all: Kalman decomposition.  
Theorem that all equivalent systems are of this form; 
and representation is unique.

@@

B and C fixed.
